Homicide Increases in the COVID-19 Era

### 2025-04-10 Draft4.qmd
Draft 4

```{r DATA, output=FALSE, echo=FALSE}
library(sf) # for map
library(tidyverse) # using only for joining datasets to build map visual
library(RColorBrewer) # for map color scheme
library(tidygeocoder) # to get zip codes of addresses
library(dplyr)

# primary datasets
# Primary datasets: Crime Data 2020-Present and 2010-2019
crime_la_2010_2019 <- read.csv("0_Data/Raw_Data/Crime_Data_from_2010_to_2019.csv")

crime_la_2020_present <- read.csv("0_Data/Raw_Data/Crime_Data_from_2020_to_Present.csv")

crime_la <- read.csv("0_Data/Cleaned_Data/all_crimes_2010_present.csv")

# primary dataset zip code and geometry data
crimes_with_zips <- read.csv("0_Data/Cleaned_Data/crimes_with_zips.csv") # just inner_join with the crime_la dataset

# Secondary dataset - Median Income by ZCTA
cali_median_income <- read.csv("0_Data/Cleaned_Data/cali_median_income.csv")

# adding a variable that's just the ZCTA zip code
cali_median_income$ZCTA <- stringr::str_sub(cali_median_income$NAME, -5)

# Secondary dataset - COVID cases in Los Angeles County
covid_14day <- read.csv("0_Data/Raw_Data/LA_County_Covid19_CSA_14day_case_death_table.csv")

evictions_la <- read.csv("0_Data/Raw_Data/LA_evictions.csv")

crime_la$DateParsed <- as.POSIXct(crime_la$Date.Rptd, 
                                  format = "%m/%d/%Y %I:%M:%S %p", 
                                  tz = "America/Los_Angeles")

crime_la$month <- lubridate::month(crime_la$DateParsed)

crime_la$YearWeek <- format(crime_la$DateParsed, "%Y-%U")

crime_la$year <- as.integer(stringr::str_sub(crime_la$YearWeek, 1, 4))
```

@@

```{r echo=FALSE, fig.cap="Crime densities in Los Angeles by LAPD Division, message=FALSE, warning=FALSE, 2010.2019 and 2020.Present", fig.align="center"}
lapd_shape <- read_sf(dsn = "0_Data/Raw_Data/LAPD_Divisions", layer = "LAPD_Divisions")

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap = "Weekly Crime Counts in Los Angeles from 2010 to present", fig.align = "center"}
crime_la$DateParsed <- as.POSIXct(crime_la$Date.Rptd, 
                                  format = "%m/%d/%Y %I:%M:%S %p", 
                                  tz = "America/Los_Angeles")

crime_la$month <- lubridate::month(crime_la$DateParsed)

crime_la$YearWeek <- format(crime_la$DateParsed, "%Y-%U")

weekly_counts <- aggregate(DR_NO ~ YearWeek, data = crime_la, FUN = length)
names(weekly_counts)[2] <- "Crimes"

weekly_counts <- weekly_counts[order(weekly_counts$YearWeek), ]

weekly_counts$Index <- seq_len(nrow(weekly_counts))

plot(
  weekly_counts$Index,
  weekly_counts$Crimes,
  type = "l",
  col  = "blue",
  lwd  = 2,
  xlab = "Weekly Index",
  ylab = "Number of Crimes")

abline(h = pretty(weekly_counts$Crimes), v = pretty(weekly_counts$Index), 
       col = "gray", lty = "dotted")
```

Figure 2 analyzes the number of crimes over time using time series. The time series shows stable weekly crime counts with occasional large spikes. These surges indicate certain weeks where crime reporting jumped significantly, there is a recent interesting surge that seems to mean-revert in 2024.

```{r, echo=FALSE, fig.cap = "Los Angeles Crimes by Day of the Week", warning=FALSE, message=FALSE, fig.align = "center"}
# Convert Date.Rptd to Date format (ignore time part)
#crime_la$Date.Rptd
crime_la$Date <- as.Date(substr(crime_la$Date.Rptd, 1, 10), format = "%m/%d/%Y")

# Extract the day of the week
crime_la$DayOfWeek <- weekdays(crime_la$Date)
# crime_la$DayOfWeek

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Percent of crime victims by gender, 2010 to present", fig.align = "center"}
# Remove missing values and empty strings in Vict.Sex column
crime_la <- subset(crime_la, Vict.Sex != "" & !is.na(Vict.Sex))

# Replace all non-"F" and non-"M" values with "N"
crime_la$Vict.Sex <- ifelse(crime_la$Vict.Sex %in% c("F", "M"), 
                                  crime_la$Vict.Sex, "N")

# Count occurrences of each gender
gender_counts <- table(crime_la$Vict.Sex)

# Create a pie chart using base R
color <- c("F" = "pink", "M" = "lightblue", "N" = "purple")
pie(
  gender_counts, 
  labels = paste(names(gender_counts), 
                 round(gender_counts / sum(gender_counts) * 100, 1), "%"), 
  col = color[names(gender_counts)]
)
```

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center"}
zcta <- read_sf(dsn = "0_Data/Raw_Data/cb_2020_us_zcta520_500k", layer = "cb_2020_us_zcta520_500k")
LA_zctas <- read.csv("0_Data/Raw_Data/zctas.csv") #obtained using zipcodeR::search_city("Los Angeles", "CA") and then checking which zip codes are also ZCTAs

@@

# getting LA median income using joined file
median_household_income <- cali_median_income %>% dplyr::select(ZCTA, S1903_C03_001E, year)

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Annual Median Income and Homicide Count by ZCTA in Los Angeles with top 1% income and top 1% homicide rates labelled", fig.align = "center"}
cali_median_income$ZCTA5CE20 <- as.integer(cali_median_income$ZCTA)
crime_income <- crimes_with_zips %>% inner_join(cali_median_income, by = c("year", "ZCTA5CE20"))

@@

crime_income_type <- crime_income %>% inner_join(crime_type, by = c("year", "DR_NO"))

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center", fig.cap="Monthly Frequency of Evictions and Crimes in the City of Los Angeles, February 2023 to February 2025"}
# fix mistake
evictions_la$Date.Received <- gsub("2205", "2025", evictions_la$Date.Received)

@@

# create color scheme
crime_la$month_year <- format(crime_la$DateParsed, "%Y-%m")
crime_la_monthly <- crime_la %>% 
  group_by(month_year) %>% 
  summarize(num_crimes = n())

@@

bureaus <- read.csv("0_Data/Raw_Data/Division_Bureau_Ref.csv")

@@

crime_by_area <- crime_by_area %>% left_join(bureaus, join_by("AREA.NAME" == "Division"))

@@

# since we do not have pre- and post- pandemic city of LA data, we are instead using LA county data for evictions. 
LA_county_evictions <- read.csv("0_Data/Raw_Data/LA_county_evictions.csv")

# covid data
covid_14day$date <- ymd(covid_14day$ep_date)
covid_14day$month <- month(covid_14day$date)
covid_14day$year <- year(covid_14day$date)
covid_14day$date <- day(covid_14day$date)
# since there is a 7 day lag in reporting, let's do monthly = cases on the 21st and on the 5th of the next month.
covid_14day$lagged <- covid_14day$date + 7

@@

# next, unemployment rate
unemployment <- read.csv("0_Data/Raw_Data/la_county_unemployment_rate.csv")
unemployment$month <- as.numeric(gsub("M", "", unemployment$Period))

# lapd budget
lapd_budget <- read.csv("0_Data/Raw_Data/lapd_budget.csv")

@@

correlations_df <- crime_la_monthly %>% inner_join(unemployment, join_by("year" == "Year", "month" == "month")) %>% dplyr::select(month, year, Value, num_crimes) %>% rename("unemployment_rate" = "Value")

@@

correlations_df4 <- correlations_df3 %>% inner_join(lapd_budget, by = "year")

correlations <- correlations_df4 %>% dplyr::select(year, month, unemployment_rate, num_crimes, cases, num_evictions, police_budget)

@@

long_Cor <- reshape2::melt(Cor_mat)

long_Cor <- long_Cor %>%
    mutate(Var1 = recode(Var1, year = 'Year', month = 'Month', unemployment_rate =  'Unemployment Rate', num_crimes = "Number of Crimes", cases = "COVID Cases", num_evictions = "Number of Evictions", police_budget = "LAPD Budget ($ billions)"),
           Var2 = recode(Var2, year = 'Year', month = 'Month', unemployment_rate =  'Unemployment Rate', num_crimes = "Number of Crimes", cases = "COVID Cases", num_evictions = "Number of Evictions", police_budget = "LAPD Budget \n ($ billions)"))

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center", fig.cap="Homicide Victims by Age and Gender in Los Angeles 2010-Present"}
homicide_by_gender <- crime_la %>% filter(Crm.Cd.Desc == "CRIMINAL HOMICIDE") %>% dplyr::select(Vict.Sex, DateParsed, Vict.Age) %>% mutate(gender = factor(Vict.Sex))

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center", fig.cap="Time series analyzing homicide rate trends over time by median income quartiles"}
# Data Preparation
homicides <- crime_la %>% inner_join(crimes_with_zips, by = c("DR_NO", "LAT", "LON","X", "year")) %>% filter(Crm.Cd.Desc == "CRIMINAL HOMICIDE") %>% 
  mutate(date = as.Date(DateParsed),
         zip = as.character(ZCTA5CE20)) %>% dplyr::select(-X)

homicides <- homicides %>% inner_join(cali_median_income, join_by("zip" == "ZCTA", "year" == "year"))
homicides$S1903_C03_001E <- as.numeric(homicides$S1903_C03_001E)

@@

@@

@@

```{r, warning=FALSE, message=FALSE, echo=FALSE, fig.align = "center"}
library(kableExtra)
crimes_by_cb <- read.csv("0_Data/Cleaned_Data/crimes_by_census_block.csv")
colnames(crimes_by_cb) <- c(colnames(crimes_by_cb)[-c(1, length(colnames(crimes_by_cb)))],"LON", "LAT")

crimes_by_cb$LON <- as.numeric(gsub('c(', "", crimes_by_cb$LON, fixed = TRUE))
crimes_by_cb$LAT <- as.numeric(gsub(')', "", crimes_by_cb$LAT, fixed = TRUE))

crimes_by_cb$DateParsed <- as.POSIXct(crimes_by_cb$Date.Rptd, 
                                  format = "%m/%d/%Y %I:%M:%S %p", 
                                  tz = "America/Los_Angeles")

crimes_by_cb$YearWeek <- format(crimes_by_cb$DateParsed, "%Y-%U")

crimes_by_cb$year <- as.integer(stringr::str_sub(crimes_by_cb$YearWeek, 1, 4))

homicides_by_cb <- crimes_by_cb %>% filter(Crm.Cd.Desc == "CRIMINAL HOMICIDE") %>% group_by(BLKGRPCE, year) %>% summarize(HomicideCount = n()) %>% ungroup()

@@

@@

#### Data Wrangling {.unnumbered}

##### Reading in Raw Datasets

```{r DATA_Wrangling, output=FALSE, eval=FALSE}
library(sf) # for map
library(tidyverse) # using only for joining datasets to build map visual
library(RColorBrewer) # for map color scheme
library(tidygeocoder) # to get zip codes of addresses
library(dplyr)

# primary datasets
# Primary datasets: Crime Data 2020-Present and 2010-2019
crime_la_2010_2019 <- read.csv("0_Data/Raw_Data/Crime_Data_from_2010_to_2019.csv")

crime_la_2020_present <- read.csv("0_Data/Raw_Data/Crime_Data_from_2020_to_Present.csv")

crime_la <- read.csv("0_Data/Cleaned_Data/all_crimes_2010_present.csv")

# primary dataset zip code and geometry data
crimes_with_zips <- read.csv("0_Data/Cleaned_Data/crimes_with_zips.csv") # just inner_join with the crime_la dataset

# Secondary dataset - Median Income by ZCTA
cali_median_income <- read.csv("0_Data/Cleaned_Data/cali_median_income.csv")

# adding a variable that's just the ZCTA zip code
cali_median_income$ZCTA <- stringr::str_sub(cali_median_income$NAME, -5)

# Secondary dataset - COVID cases in Los Angeles County
covid_14day <- read.csv("0_Data/Raw_Data/LA_County_Covid19_CSA_14day_case_death_table.csv")

evictions_la <- read.csv("0_Data/Raw_Data/LA_evictions.csv")

crime_la$DateParsed <- as.POSIXct(crime_la$Date.Rptd, 
                                  format = "%m/%d/%Y %I:%M:%S %p", 
                                  tz = "America/Los_Angeles")

crime_la$month <- lubridate::month(crime_la$DateParsed)

crime_la$YearWeek <- format(crime_la$DateParsed, "%Y-%U")

crime_la$year <- as.integer(stringr::str_sub(crime_la$YearWeek, 1, 4))
```

```{r Combining Median Income, eval=FALSE}
# Secondary datasets
california_median_income_2018_2023 <- list()
yrs <- 2018:2023
for (i in 1:length(yrs)) {
  obj <- read.csv(paste0("0_Data/Raw_Data/ACSST5Y",yrs[[i]],".S1903-Data.csv"), skip = 0)
  obj <- obj[-1,]
  obj$year <- yrs[[i]]
  california_median_income_2018_2023[[i]] <- obj
}

cali_median_income <- do.call(rbind, california_median_income_2018_2023)

# adding a variable that's just the ZCTA zip code
cali_median_income$ZCTA <- stringr::str_sub(cali_median_income$NAME, -5)

write.csv(cali_median_income, "cali_median_income.csv")
```

##### Zip Codes for Crimes

```{r Get Zip Codes, eval=FALSE}
library(sf)
library(dplyr)

crime_la_2010_2019 <- read.csv("0_Data/Raw_Data/Crime_Data_from_2010_to_2019.csv")

crime_la_2020_present <- read.csv("0_Data/Raw_Data/Crime_Data_from_2020_to_Present.csv")

# Combine the two datasets
crime_la <- rbind(crime_la_2010_2019, crime_la_2020_present)

crime_la$DateParsed <- as.POSIXct(crime_la$Date.Rptd, 
                                  format = "%m/%d/%Y %I:%M:%S %p", 
                                  tz = "America/Los_Angeles")

crime_la$YearWeek <- format(crime_la$DateParsed, "%Y-%U")

crime_la$year <- as.integer(stringr::str_sub(crime_la$YearWeek, 1, 4))

coords <- crime_la %>% dplyr::select(LAT, LON, DR_NO, year)

zcta <- read_sf(dsn = "0_Data/Raw_Data/cb_2020_us_zcta520_500k", layer = "cb_2020_us_zcta520_500k")

  points <- st_as_sf(coords, coords = c("LON", "LAT"), crs = 4326)  # WGS84

  zcta <- st_transform(zcta, 4326)  # Convert ZCTA to WGS84 if needed
  cat("Starting Zip Code Retrieval \n")

  result <- st_join(points, zcta)

  cat("Finished Zip Code Retrieval \n")

  write.csv(result, "0_Data/Raw_Data/all_zips.csv")

crime_zctas <- read.csv("0_Data/Raw_Data/all_zips.csv")
colnames(crime_zctas) <- c(colnames(crime_zctas)[-c(1, length(colnames(crime_zctas)))],"LON", "LAT")

crime_zctas$LON <- as.numeric(gsub('c(', "", crime_zctas$LON, fixed = TRUE))
crime_zctas$LAT <- as.numeric(gsub(')', "", crime_zctas$LAT, fixed = TRUE))

# Repeat same process for Census Blocks
```

```{r Get Zip Codes 2, eval=FALSE}

duplicates <- as.integer(as.character(data.frame(table(crime_zctas$universal_idx)[table(crime_zctas$universal_idx) > 1]) %>% pull(Var1)))

# there are approximately 200 rows that are duplicated (100 incidents) as the area lies on the intersection between two zctas. 
duplicate_data <- crime_zctas %>% filter(universal_idx %in% duplicates)

rev_geo_duplicates <- reverse_geocode(duplicates2, lat = LAT, long = LON)
# the above query identifies the address of these coordinates as KFC, 8644, Balboa Boulevard, Northridge South Neighborhood Council District, Los Angeles, Los Angeles County, California, 91325, United States. 
# therefore, we will only keep the 91325 zip code rows.

crime_zctas_no_duplicates <- crime_zctas %>% mutate(duplicate = ifelse(universal_idx %in% duplicate_data$universal_idx & ZCTA5CE20 != 91325, "Drop", "Keep")) %>% filter(duplicate != "Drop") %>% dplyr::select(-duplicate)

# it's empty now 
# duplicates <- table(crime_zctas_no_duplicates$universal_idx)[table(crime_zctas_no_duplicates$universal_idx) > 1]

# next, check for NAs
sum(is.na(crime_zctas_no_duplicates$ZCTA5CE20)) # there are 3724 NAs. Let's find them a zip code

# some also have lat long = 0. We'll find those separately
no_zipcode_not0 <- crime_zctas_no_duplicates %>% filter(is.na(ZCTA5CE20) & LON != 0)

no_zipcode_not0_zips <- reverse_geocode(no_zipcode_not0, lat = LAT, long = LON)

no_zipcode_not0_zips$ZCTA5CE20 <- stringr::str_extract(no_zipcode_not0_zips$address, "\\d{5}")

no_zipcode_not0_zips <- no_zipcode_not0_zips %>% dplyr::select(-address)

crime_zctas2 <- crime_zctas_no_duplicates %>% filter(!is.na(ZCTA5CE20))

crime_zctas3 <- rbind(crime_zctas2, no_zipcode_not0_zips)

crime_zctas4 <- crime_zctas3 %>% filter(!is.na(ZCTA5CE20))

write.csv(crime_zctas4, "crimes_with_zips.csv")

no_zipcode_0s <- crime_zctas_no_duplicates %>% filter(is.na(ZCTA5CE20) & LON == 0) %>% pull(DR_NO)

main_dataset_0s <- crime_la %>% filter(DR_NO %in% no_zipcode_0s)

summary(main_dataset_0s)

summary(crime_la)
```

```{r eval=FALSE, fig.cap="Crime densities in Los Angeles by LAPD Division, message=FALSE, warning=FALSE, 2010.2019 and 2020.Present", fig.align="center"}
lapd_shape <- read_sf(dsn = "0_Data/Raw_Data/LAPD_Divisions", layer = "LAPD_Divisions")

@@

```{r, eval=FALSE, warning=FALSE, message=FALSE, fig.cap = "Weekly Crime Counts in Los Angeles from 2010 to present", fig.align = "center"}
crime_la$DateParsed <- as.POSIXct(crime_la$Date.Rptd, 
                                  format = "%m/%d/%Y %I:%M:%S %p", 
                                  tz = "America/Los_Angeles")

crime_la$month <- lubridate::month(crime_la$DateParsed)

crime_la$YearWeek <- format(crime_la$DateParsed, "%Y-%U")

weekly_counts <- aggregate(DR_NO ~ YearWeek, data = crime_la, FUN = length)
names(weekly_counts)[2] <- "Crimes"

weekly_counts <- weekly_counts[order(weekly_counts$YearWeek), ]

weekly_counts$Index <- seq_len(nrow(weekly_counts))

plot(
  weekly_counts$Index,
  weekly_counts$Crimes,
  type = "l",
  col  = "blue",
  lwd  = 2,
  xlab = "Weekly Index",
  ylab = "Number of Crimes")

abline(h = pretty(weekly_counts$Crimes), v = pretty(weekly_counts$Index), 
       col = "gray", lty = "dotted")
```

>>>>

```{r DATA, output=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
library(sf) # for map
library(tidyverse) # using only for joining datasets to build map visual
library(RColorBrewer) # for map color scheme
library(tidygeocoder) # to get zip codes of addresses
library(dplyr)
library(RSQLite)

con <- dbConnect(SQLite(), "0_Data/crime_in_la.db")
```

@@

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align="center", fig.cap="Crime densities in Los Angeles by LAPD Division, 2010-2019 and 2020-Present"}
query <- dbSendQuery(con, "Select * From crime_la Where year < 2020")
crime_la_2010_2019 <- dbFetch(query,n=-1)
dbClearResult(query)

query <- dbSendQuery(con, "Select * From crime_la Where year > 2019")
crime_la_2020_present <- dbFetch(query,n=-1)
dbClearResult(query)

lapd_shape <- read_sf(dsn = "0_Data/Raw_Data/LAPD_Divisions", layer = "LAPD_Divisions")

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap = "Weekly Crime Counts in Los Angeles from 2010 to present", fig.align = "center"}
query <- dbSendQuery(con, "Select Count(*) as Crimes, YearWeek From crime_la Group By YearWeek Order By YearWeek")
weekly_counts <- dbFetch(query,n=-1)
dbClearResult(query)

weekly_counts$Index <- seq_len(nrow(weekly_counts))
# Find the Index for March 2020
march_2020_index <- which(weekly_counts$YearWeek == "2020-09")  # Adjust '202009' to match your data's YearWeek format for March 2020

# Plot
plot(
  weekly_counts$Index,
  weekly_counts$Crimes,
  type = "l",
  col  = "blue",
  lwd  = 2,
  xlab = "Weekly Index",
  ylab = "Number of Crimes")

abline(h = pretty(weekly_counts$Crimes), v = pretty(weekly_counts$Index), 
       col = "gray", lty = "dotted")

# Add vertical line for March 2020
abline(v = weekly_counts$Index[march_2020_index], col = "red", lwd = 2)

# Add a label for March 2020
text(
  x = weekly_counts$Index[march_2020_index] - 80, 
  y = 10, 
  labels = "March 2020", 
  pos = 3, 
  col = "red"
)
```

Figure 2 analyzes the number of crimes over time using time series. The time series shows stable weekly crime counts with occasional large spikes, likely due to shorter weeks. These surges indicate certain weeks where crime reporting jumped significantly, there is a recent interesting surge that seems to mean-revert in 2024. The red line, which denotes March 2020, shows a decline in crimes, likely during the shutdown and quarantine, but it returned to normal levels quickly.

```{r, echo=FALSE, fig.cap = "Los Angeles Crimes by Day of the Week", warning=FALSE, message=FALSE, fig.align = "center"}
# Convert Date.Rptd to Date format (ignore time part)
#crime_la$Date.Rptd
crime_la <- rbind(crime_la_2010_2019, crime_la_2020_present)
# Extract the day of the week
crime_la$Date <- as.Date(substr(crime_la$Date.Rptd, 1, 10), format = "%m/%d/%Y")

crime_la$DayOfWeek <- weekdays(crime_la$Date)
# crime_la$DayOfWeek

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Percent of crime victims by gender, 2010 to present", fig.align = "center"}
query <- dbSendQuery(con, "SELECT COUNT(*) As count, Modified_Vict_Sex As `Vict.Sex` FROM (
    SELECT *,  
        CASE 
            WHEN `Vict.Sex` NOT IN ('M', 'F') THEN 'N'
            ELSE `Vict.Sex`
        END AS `Modified_Vict_Sex`
    FROM crime_la c
    WHERE `Vict.Sex` != '' AND `Vict.Sex` IS NOT NULL) s
    GROUP BY Modified_Vict_Sex
")
gender_counts <- dbFetch(query,n=-1)
dbClearResult(query)

# Create a pie chart using base R
color <- c("F" = "pink", "M" = "lightblue", "N" = "purple")
pie(
  gender_counts$count, 
  labels = paste(gender_counts$Vict.Sex, 
                 round(gender_counts$count / sum(gender_counts$count) * 100, 1), "%"), 
  col = color[gender_counts$Vict.Sex]
)
```

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center"}
zcta <- read_sf(dsn = "0_Data/Raw_Data/cb_2020_us_zcta520_500k", layer = "cb_2020_us_zcta520_500k")
query <- dbSendQuery(con, "SELECT * FROM zctas_in_la")
LA_zctas <- dbFetch(query,n=-1)
dbClearResult(query)

@@

# getting LA median income using joined file
query <- dbSendQuery(con, "SELECT ZCTA, S1903_C03_001E, year FROM cali_median_income")
median_household_income <- dbFetch(query,n=-1)
dbClearResult(query)

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Annual Median Income and Homicide Count by ZCTA in Los Angeles with top 1% income and top 1% homicide rates labelled", fig.align = "center"}
zcta <- read_sf(dsn = "0_Data/Raw_Data/cb_2020_us_zcta520_500k", layer = "cb_2020_us_zcta520_500k")
query <- dbSendQuery(con, "SELECT i.*, c.*
FROM cali_median_income i
INNER JOIN crimes_with_zips c
ON i.year = c.year AND CAST(i.ZCTA AS int) = c.ZCTA5CE20")
crime_income <- dbFetch(query,n=-1)
dbClearResult(query)

@@

crime_income_type <- crime_income[,-c(249, 247)] %>% inner_join(crime_type, join_by("year" == "year", "DR_NO" == "DR_NO"))

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center", fig.cap="Monthly Frequency of Evictions and Crimes in the City of Los Angeles, February 2023 to February 2025"}

query <- dbSendQuery(con, "SELECT * FROM evictions_city_la")
evictions_la <- dbFetch(query,n=-1)
dbClearResult(query)
# fix mistake
evictions_la$Date.Received <- gsub("2205", "2025", evictions_la$Date.Received)

@@

# create color scheme
crime_la$month_year <- format(crime_la$Date, "%Y-%m")
crime_la_monthly <- crime_la %>% 
  group_by(month_year) %>% 
  summarize(num_crimes = n())

@@

query <- dbSendQuery(con, "SELECT * FROM lapd_bureaus")
bureaus <- dbFetch(query,n=-1)
dbClearResult(query)

@@

query <- dbSendQuery(con, "SELECT COUNT(*) FROM crime_la GROUP BY `AREA.NAME`, year")
test <- dbFetch(query,)
crime_by_area <- crime_by_area %>% left_join(bureaus, join_by("AREA.NAME" == "Division"))

@@

# since we do not have pre- and post- pandemic city of LA data, we are instead using LA county data for evictions.
# covid data
query <- dbSendQuery(con, "
  SELECT 
    CAST(strftime('%Y', ep_date) AS INTEGER) AS year,
    CAST(strftime('%m', ep_date) AS INTEGER) AS month,
   CAST(strftime('%d', ep_date) AS INTEGER) AS date, 
   cases_14day
  FROM covid_cases
")
covid_14day <- dbFetch(query,n=-1)
dbClearResult(query)
# since there is a 7 day lag in reporting, let's do monthly = cases on the 21st and on the 5th of the next month.
covid_14day$lagged <- covid_14day$date + 7

@@

# next, unemployment rate, lapd budget, and evictions
query <- dbSendQuery(con, "
  SELECT 
    u.month, 
    u.Year, 
    u.Value AS unemployment_rate, 
    l.police_budget, 
    e.Count AS num_evictions
  FROM 
    unemployment_la AS u
  INNER JOIN 
    lapd_budget AS l ON u.Year = l.year
  INNER JOIN 
    evictions_county_la AS e ON u.Year = e.Year AND u.month = e.Month
")
unemployment_lapd_budget_evictions <- dbFetch(query,n=-1)
dbClearResult(query)

@@

correlations_df <- crime_la_monthly %>% inner_join(unemployment_lapd_budget_evictions, join_by("year" == "Year", "month" == "month")) %>% dplyr::select(month, year, num_evictions, unemployment_rate, num_crimes, police_budget)

@@

correlations <- correlations_df2 %>% dplyr::select(year, month, unemployment_rate, num_crimes, cases, num_evictions, police_budget)

@@

# get only the upper triangle of variables 
# (because a correlation heat map shows each correlation value twice)
get_upper_tri <- function(cormat){
  # lower.tri -> Returns a matrix of logicals, so set the values that are true to NA
  cormat[lower.tri(cormat)]<- NA # set the correlation matrix lower triangle to NA
  return(cormat) 
}

# reorder them by the strength of the correlation
reorder_cormat <- function(cormat){
  # Use correlation between variables as distance
  dd <- as.dist((1-cormat)/2)
  hc <- hclust(dd) # use hierarchical clustering to reorder
  cormat <-cormat[hc$order, hc$order]
}

# use the functions above to remove the unnecessary parts
Cor_mat<-reorder_cormat(Cor_mat)

Cor_mat_upper<-get_upper_tri(Cor_mat)

# using melt instead of pivot_longer because pivot_longer did not order correctly
Cor_mat_upper<-reshape2::melt(Cor_mat_upper, na.rm=TRUE)

# remove the central values 
# which show the correlation for each variable with itself
Cor_mat_upper<- Cor_mat_upper %>% filter(value != 1.00)

long_Cor <- Cor_mat_upper %>%
    mutate(Var1 = recode(Var1, year = 'Year', month = 'Month', unemployment_rate =  'Unemployment Rate', num_crimes = "Number of Crimes", cases = "COVID Cases", num_evictions = "Number of Evictions", police_budget = "LAPD Budget ($ billions)"),
           Var2 = recode(Var2, year = 'Year', month = 'Month', unemployment_rate =  'Unemployment Rate', num_crimes = "Number of Crimes", cases = "COVID Cases", num_evictions = "Number of Evictions", police_budget = "LAPD Budget \n ($ billions)"))

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center", fig.cap="Homicide Victims by Age and Gender in Los Angeles 2010-Present"}
homicide_by_gender <- crime_la %>% filter(Crm.Cd.Desc == "CRIMINAL HOMICIDE") %>% dplyr::select(Vict.Sex, DateParsed, Vict.Age) %>% mutate(gender = factor(Vict.Sex, levels  = c("F", "M", "X"), labels = c("Female", "Male", "Other"))) %>% drop_na()

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center", fig.cap="Time series analyzing homicide rate trends over time by median income quartiles"}
query <- dbSendQuery(con, "
  SELECT i.S1903_C03_001E, x.year, i.ZCTA, x.LAT, x.LON
  FROM (
    SELECT c.LAT, c.LON, CAST(z.ZCTA5CE20 AS INTEGER) AS ZCTA5CE20, CAST(c.year AS INTEGER) AS year
    FROM crime_la AS c
    INNER JOIN crimes_with_zips AS z 
      ON c.DR_NO = z.DR_NO 
      AND c.LAT = z.LAT 
      AND c.LON = z.LON 
      AND c.X = z.X 
      AND c.year = z.year
    WHERE c.\"Crm.Cd.Desc\" = 'CRIMINAL HOMICIDE'
  ) AS x INNER JOIN (SELECT m.S1903_C03_001E, CAST(m.year AS INTEGER) AS year, CAST(m.ZCTA AS INTEGER) AS ZCTA FROM cali_median_income m) AS i 
    ON x.ZCTA5CE20 = i.ZCTA AND x.year = i.year
")

homicides <- dbFetch(query,n=-1)
dbClearResult(query)

@@

\begin{equation}
\text{Crimes}_i = \beta_0 + \beta_1 \text{Index}_i + \beta_2 \text{CrimesLag}_i + \epsilon_i
\end{equation}

@@

\begin{equation}
\text{HomicideCount}_i = \beta_0 + \beta_1 \text{cali\_median\_income}_i + \epsilon_i
\end{equation}

@@

\begin{equation}
\text{HomicideCount}_i = \beta_0 + \sum_{j=1}^{k} \beta_j \text{Factor}_j + \epsilon_i
\end{equation}

```{r, warning=FALSE, message=FALSE, echo=FALSE, fig.align = "center"}
library(kableExtra)
query <- dbSendQuery(con, "SELECT * FROM crimes_by_census_block")
crimes_by_cb <- dbFetch(query,n=-1)
dbClearResult(query)

homicides_by_cb <- crimes_by_cb %>% filter(str_detect(Crm.Cd.Desc, "CRIMINAL HOMICIDE")) %>% group_by(BLKGRPCE, year) %>% summarize(HomicideCount = n()) %>% ungroup()

@@

Model 1 is defined as: 
\begin{equation}
\text{HomicideCount}_i = \beta_0 + \beta_1 \text{LAT\_rounded}_i + \beta_2 \text{LON\_rounded}_i + \epsilon_i
\end{equation}

Model 2 is defined as: 
\begin{equation}
\text{HomicideCount}_i = \beta_0 + \beta_1 \text{year}_i + \epsilon_i
\end{equation}

Model 3 is defined as:
\begin{equation}
\text{HomicideCount}_i = \beta_0 + \beta_1 \text{LAT\_rounded}_i + \beta_2 \text{LON\_rounded}_i + \beta_3 \text{year}_i + \epsilon_i
\end{equation}

Model 4 is defined as:
\begin{align}
\text{HomicideCount}_i &= \beta_0 + \beta_1 \text{LAT\_rounded}_i + \beta_2 \text{LON\_rounded}_i + \beta_3 \text{year}_i \nonumber \\
&\quad + \beta_4 (\text{LAT\_rounded}_i \times \text{LON\_rounded}_i) + \beta_5 (\text{LAT\_rounded}_i \times \text{year}_i) \nonumber \\
&\quad + \beta_6 (\text{LON\_rounded}_i \times \text{year}_i) + \beta_7 (\text{LAT\_rounded}_i \times \text{LON\_rounded}_i \times \text{year}_i) + \epsilon_i
\end{align}

@@

```{r}
knitr::opts_chunk$set(eval = FALSE)
```

```{r, output=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
library(sf) # for map
library(tidyverse) # using only for joining datasets to build map visual
library(RColorBrewer) # for map color scheme
library(tidygeocoder) # to get zip codes of addresses
library(dplyr)
library(RSQLite)

con <- dbConnect(SQLite(), "0_Data/crime_in_la.db")
```

```{r echo=FALSE, fig.cap="Crime densities in Los Angeles by LAPD Division, message=FALSE, warning=FALSE, 2010.2019 and 2020.Present", fig.align="center"}
query <- dbSendQuery(con, "Select * From crime_la Where year < 2020")
crime_la_2010_2019 <- dbFetch(query,n=-1)
dbClearResult(query)

query <- dbSendQuery(con, "Select * From crime_la Where year > 2019")
crime_la_2020_present <- dbFetch(query,n=-1)
dbClearResult(query)

lapd_shape <- read_sf(dsn = "0_Data/Raw_Data/LAPD_Divisions", layer = "LAPD_Divisions")

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap = "Weekly Crime Counts in Los Angeles from 2010 to present", fig.align = "center"}
query <- dbSendQuery(con, "Select Count(*) as Crimes, YearWeek From crime_la Group By YearWeek Order By YearWeek")
weekly_counts <- dbFetch(query,n=-1)
dbClearResult(query)

weekly_counts$Index <- seq_len(nrow(weekly_counts))
# Find the Index for March 2020
march_2020_index <- which(weekly_counts$YearWeek == "2020-09")  # Adjust '202009' to match your data's YearWeek format for March 2020

# Plot
plot(
  weekly_counts$Index,
  weekly_counts$Crimes,
  type = "l",
  col  = "blue",
  lwd  = 2,
  xlab = "Weekly Index",
  ylab = "Number of Crimes")

abline(h = pretty(weekly_counts$Crimes), v = pretty(weekly_counts$Index), 
       col = "gray", lty = "dotted")

# Add vertical line for March 2020
abline(v = weekly_counts$Index[march_2020_index], col = "red", lwd = 2)
